A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-09-11 jour/jour.tex
more

@@

\makeatletter
\def\spnXwtheorem#1#2#3#4{\@spynthm{#1}{#2}{#3}{#4}%
                         \@addtoreset{#1}{chapter}}%
\makeatother
\spnXwtheorem{examplex}{Example}{\itshape}{\rmfamily}

@@

\makeatletter
\ps@myheadings
\makeatother

\include{defs}

@@

\author {Andrew Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3}}
\authorrunning {A. Reynolds \and J. C. Blanchette}
\institute{
\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Inria Nancy \& LORIA, Villers-l\`es-Nancy, France
\and
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
}

@@

@@

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. As a simple example, the coalgebraic specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
%(using an ML-like syntax)
introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
in Peano notation but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies
%the value~
$\infty$.
Compared with the conventional %algebraic
definition
\[\vthinspace \keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
%with a dedicated constructor for representing infinity, the
the codatatype avoids
one case by unifying the infinite and finite nonzero cases.

@@

{\footnotesize
\def\ackname{Acknowledgment}
\paragraph{\footnotesize\ackname.}
\looseness=-1
We owe a great debt to the development team of CVC4,
including Clark Barrett and Cesare Tinelli,
and in particular Morgan Deters,
who jointly with the first author
developed the initial version
of the theory solver for datatypes in CVC4.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatypes or at least benevolently tolerated it, both of which we are thankful
for.
%
Peter Gammie and Andreas Lochbihler provided useful benchmarks.
Andrei Popescu helped clarify our thoughts regarding codatatypes and indicated
related work. Dmitriy Traytel took part in discussions about degenerate
codatatypes.
%
Pascal Fontaine, Andreas Lochbihler, Andrei Popescu, Christophe Ringeissen, Mark
Summerfield, Dmitriy Traytel, and the anonymous reviewers suggested many
textual improvements. The second author's research was partially supported by
the Deutsche
Forschungs\-gemein\-schaft %(DFG)
project ``Den Hammer h\"arten'' (grant NI\,491\slash 14-1)
and the Inria technological development
action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC).

@@

\bibliographystyle{splncs03}
\bibliography{bib}{}

>>>>

%%% GOOD IDEAS in impl of codata models:
%%% 1. de Bruijn
%%% 2. enum: "X nat"
%%% 3. norm by DFA minim
%%% 4. check membership in bad set by matching

@@

\spnewtheorem{examplex}{Example}{\itshape}{\rmfamily}

@@

@@

\author {Andrew Reynolds \and Jasmin Christian Blanchette}
\authorrunning {A. Reynolds \and J. C. Blanchette}

\institute{
Andrew Reynolds
\at \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Jasmin Christian Blanchette%, funded by DFG grant \textit{Hardening the Hammer} (Ni\,491\slash 14-1)
\at Inria \& LORIA, Nancy, France \\
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany}

@@

\bigskip

@@

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. As a simple example, the coalgebraic specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
%(using an ML-like syntax)
introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
in Peano notation but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies
%the value~
$\infty$.

@@

{\footnotesize
\def\ackname{Acknowledgment}
\paragraph{\footnotesize\ackname.}
\looseness=-1
We owe a great debt to the development team of CVC4,
including Clark Barrett and Cesare Tinelli,
and in particular Morgan Deters,
who jointly with the first author
developed the initial version
of the theory solver for datatypes in CVC4.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatypes or at least benevolently tolerated it, both of which we are thankful
for.
%
Peter Gammie and Andreas Lochbihler provided useful benchmarks.
Andrei Popescu helped clarify our thoughts regarding codatatypes and indicated
related work. Dmitriy Traytel took part in discussions about degenerate
codatatypes.
%
Pascal Fontaine, Andreas Lochbihler, Andrei Popescu, Christophe Ringeissen, Mark
Summerfield, Dmitriy Traytel, and the anonymous reviewers suggested many
textual improvements. The second author's research was partially supported by
the Deutsche
Forschungs\-gemein\-schaft %(DFG)
project ``Den Hammer h\"arten'' (grant NI\,491\slash 14-1)
and the Inria technological development
action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC).

@@

\bibliographystyle{spmpsci}
\bibliography{bib}{}
